On the Synchronization Power of the Openflow Data Plane

### 2015-05-21 inband-sync.tex
one more partial update

@@

\begin{abstract}
Control planes of modern Software-Defined Networks (SDNs)
are \emph{distributed systems}: to ensure availability and fault-tolerance,
to improve load-balancing, and to reduce overheads,
control modules are physically distributed.
%a redundant control plane
%can ensure a high availability, and a spatially distributed
%control plane allows to handle time-critical dataplane
%events close to their origins.
%[[PK this is a contraversial point, not important for us
%Moreover, an SDN may be managed by multiple administrators simultaneously.
%]]
However, a distributed control plane also introduces
new challenges, and in order to guarantee a consistent network operation,
the actions of different controllers may need to be synchronized.
%[[PK sounds contradicting to the fact that such systems already exist
%However, not much is known today about how to implement
%such a distributed control plane.
%]]
%
A well-known concept to solve synchronization and consistency problems
in distributed systems are \emph{transactions}: a transaction defines a sequence
of operations which are performed in an atomic, serializable manner, providing all-or-nothing semantics.
This paper shows that powerful transactions can also be implemented in software-defined networks,
using the standard Openflow protocol.
%Indeed, recent Openflow versions already come with useful but primitive
%features, in particular \emph{bundling} and \emph{rule-overlap checks}.
In particular, we present a
transactional interface which supports arbitrary atomic
transactions
over the switch configuration space: Our transactions
can include complex notions of conflicts (beyond flow overlap checks),
as well as computations and multiple controller interactions.
We discuss applications of our transactional memory interface to
in-band controller consensus, consistent policy installation, and load-balancing.
\end{abstract}

@@

While the logically centralized perspective offered by SDN is attractive,
there is a wide consensus that
the control plane should be physically \emph{distributed}.
First, in order to provide a high availability as well as a minimal degree of
fault-tolerance, controllers should be redundant~\cite{onix,stn,onos}: a failure
of one controller can be masked by other controllers. Second, it has also been proposed
to distribute controllers \emph{spatially}, in order to handle latency-sensitive and
communication intensive control plane events close to their origin.~\cite{devoflow,kandoo,jukka,disco}
Third, larger SDNs are likely to be operated by multiple administrators or may even offer
participatory interfaces where different users can install and trigger policy changes
concurrently~\cite{participatory,stn}.

@@

\section{Background}\label{sec:background}
\petr{ the 1st sentence does not belong here}\liron{I think it serves as a motivation for the section, but it fines by me to remove it}
The transactional interface presented in this paper can be implemented in
Openflow~\cite{of-spec}, the standard SDN protocol today.
% In this section,
%we provide the necessary background.
In a nutshell,
Openflow
follows a match-action concept: Openflow switches store
rules (installed by the controller) consisting of a match and an
action part. A packet matched by a certain rule will be subject
to the associated action (to be immediately applied now or later).\petr{"to be immediately applied now or later" - sounds funny}\liron{I agree it sounds a bit strange, maybe we should explain this later after we disscuss the pipeline}
For example, an action can define a port to which the
matched packet should be forwarded, or add or change a tag
(a certain part in the packet header).
Each Openflow switch stores one or multiple flow tables,
each of which contains a set of rules: the \emph{flow entries}.

Flow
tables form a pipeline, and flow entries are ordered according
to priorities: A packet arriving at a switch is first checked by
the rule of the highest priority in table 0: the fields of the
data packet are compared with the match fields of that rule,
and if they fit, some instructions (the actions)  are executed; \petr{ what is the difference between instruction or action?} \liron{we should explain that: every packet is associated with an action set. This set is empty at the beginning and is changed according to the flow table matches. Each match is asociated with instructions that can add or modify actions in the action set and also to apply (now) all actions in the action set. By default action set is applied at the end of the pipeline.}
subsequently, lower priority rules are checked. Depending on
the outcome of the table 0 processing, the packet may be sent
to additional flow tables in the pipeline.
Instructions can be
used to define additional tables to be visited (a ``goto instruction''),
to modify the set of actions to be applied later (either by appending,
deleting, or modifying actions), or to immediately apply some
actions to the packet.

During the traversal of
the flow tables, the packet can be equipped with meta-tags to
transfer temporary information between tables.
The meta-data field is part of the header that can be
inserted or removed from a packet via push and pop actions.
In general, a packet can be matched against any of its header
fields, and fields can be wildcarded and sometimes bitmasked.
(For instance, the meta-data field is maskable.) If no rule
matches, the packet is dropped.

@@

\paragraph{The FlowMod Command}
The main messages used in Openflow
to update the switch configuration is called \emph{FlowMod}.
In a nutshell, \emph{FlowMod} either specifies a new flow entry or
modifications to existing flow entries.
A flow entry consists of: (1) a match defined by a ternary pattern over packet header fields
(a value and a mask), (2) a set of actions, and (3) a priority.
%,a match and an action.
The standard behavior is that the new flow entry will simply \emph{replace}
an existing one if the existing one has exactly the \emph{same match},
but will be installed
\emph{in addition} to existing entries if they only
partially overlap (or define flowspace subsets or supersets).
The latter can result
in inconsistencies, especially if the old and the new rule
have the same priority.

In order to avoid conflicts, \emph{FlowMod} can be used together with a the $\checko$ flag:
if the flow spaces of an old
and the new rule overlap,
that is, if there is an entry with an overlapping match part but a
different action,
 and if the two rules have the same priority, then
\emph{FlowMod} will fail.

We will exploit the \emph{FlowMod} command in combination with
$\checko$ to implement synchronization objects which form
the basis of our transactional interface.
We will use the following notation to create a \emph{FlowMod} message:
$\FlowMod(\match, \action, \flags)$, $\match$ denotes the match,
$\action$ the action, and $\flags$ the flags to activate the $\checko$ feature.
The \emph{FlowMod} message (and also the flow entries) include more parameters,
which however are irrelevant for our implementation: we suggest to
use the default values.

@@

Another interesting feature in Openflow
is the \emph{bundle}. A bundle allows to send multiple commands (e.g., \emph{FlowMod} messages) with the same bundle id and then execute them  by sending a bundle commit message. Theses (bundled) commands are performed with \emph{all-or-nothing semantics}:
either all of the commands are
performed, or none of them is. In particular, if, for some reason, the
configuration request contained in one of the bundled commands cannot
be performed, all the other requests  are rejected and an error
message is sent to the controller that issued them.
When properly configured,
the bundled processing can be done in a
\emph{packet-atomic} way, in the sense that
each packet is processed according
to the switch configuration either \emph{before} or \emph{after} the
bundle (i.e., all its commands) took effect. \liron{we need to disscuss controller atomicity and order preservation inside a bundle}
If multiple \emph{FlowMod} messages with the $\checko$ flag are sent as a single bundle,
the bundle command also returns a reason
why the overlap checks failed: it returns the failed command in the bundle.
Concretely, when the commit fails, the switch must generate an error message corresponding to the
message that could not be applied. The error message must use the identifier (\emph{xid}) of the offending message,
the error data field corresponding to that message and the error code corresponding to the failure. If
multiple messages are generating errors, the switch may return only the first error found or generate
multiple error messages for the same bundle.

@@

In this paper, we will use the notation
%$\textbf{bundle}\{\;\ldots\;\}\rightarrow res$ to refer to represent a
$\textbf{bundle}\{\;\ldots\;\}$ to refer to a
sequence of commands sent in a bundle. The bundle returns
$\top$ if and only if
all commands are performed successfully, and otherwise $\bot$.
\petr{What about replacing $\top-\bot$ with ack-nack? Can we also get a reason why a bundle did not commit?}
The bundles used in this paper will always have the following structure:
\begin{enumerate}
\item First, it contains a \emph{create bundle} message: a \texttt{ofpt\_bundle\_control} message of type
\texttt{ofpbct\_open\_request}.
\item The follows a sequence of commands (e.g., \emph{FlowMod}), each wrapped in an
\emph{add to bundle message} (\texttt{ofpt\_bundle\_add\_message}) with the \emph{bundle identifier}.
\item Finally, a \emph{commit bundle} control message (\texttt{ofpt\_bundle\_control}) of
type \texttt{ofpbct\_commit\_request}.
\end{enumerate}
Again, the bundle always returns a binary result: \emph{error} ($\bot$) or \emph{success} ($\top$).

@@

\paragraph{Remarks}
As we will see, in our implementation,
the switch \emph{configuration} consists of two parts:
the ``normal'' configuration which is defined by the network policy and
directly effects the traffic (e.g., forwarding rules), and some additional
configuration which serves as synchronization information 
for our mechanism. 
By installing Openflow \emph{flow monitors} with the \textsf{OFPFMF\_INITIAL} optional flag,  
a controller can read the current configuration. 
% %(and subscribe for updates). 
%
%can i match multiple flows in multiple tables? yes, but only for FlowMod, not GroupMod: but that is good enough!%
\liron{I changed a bit the description of flow monitors following Petr's comment, I hope it's more clear now.}

We also remark that our solution is only based on flow tables
and does not make use
of \emph{Openflow group tables}, nor do we make use of \emph{cookies}: rules
are always identified by exact matches.

\section{Limitations and Motivation}\label{sec:motivation}

 generic way for multiple distributed controllers to safely update switch configuration with no need of any control plane

At first sight, the Openflow bundle feature seems to provide everything
we need to implement our transactional interface \petr{did we say what we need already?}: the bundle
 defines a sequence of read and write operations
 on the switch state which is executed atomically.
 However, while the bundle can indeed be seen
 as a ``mini-transaction'', it comes with several
 limitations, which we discuss in the following.
 \petr{unclear what "reads and writes" are here, not discussed before. How openflow can be used to implement reads and writes?}

\liron{following Petr's comments, I suggest the following paragraph}
At first sight, the Openflow bundle feature seems to provide a simple and generic way for a controller to atomically update switch configuration, similarly to a transaction in shared memory system.
However, while the bundle can indeed be seen
 as a ``mini-transaction'', it comes with several
 limitations, which we discuss in the following.

\liron{ Also updated the following paragraph:}
First, if employed without any additional features, a bundle does not allow
 to change the switch state in any non-trivial way:
the commands in the bundle essentially do
 not support any conditional \emph{modifications} or computations which depend on any current switch configuration (except to limited types of constraints, e.g., free space).

But also bundling FlowMod commands with the check overlap feature ($\checko$)
comes with limitations: while it is possible
 to \emph{abort} a bundle depending on certain conditions
 in the \emph{overlapping
 flow spaces}, oftentimes it is desirable to be able
 to define more general conditions and notions of conflicts
 under which an atomic update should not take effect.
 \liron{replaced transaction in the last sentence with " atomic update"}

  Moreover,
 sometimes it is desirable to react to conflicts in smarter ways than
 to simply abort, e.g.,
 by supporting \emph{modifications}.
Let us consider two examples.

\begin{enumerate}
\item \textbf{Example 1: More complex dependencies.} Let us consider two controllers
in charge
of \emph{load-balancing}.
In a load-balancing application, seemingly independent actions, defined over
completely independent logical flow spaces (say, two flows with source IP address ranges $1*$ and $0*$),
may in fact be dependent due to collocation in the underlying physical network.
If the two controllers concurrently and independently
update forwarding rules,
they may unvoluntarily
unbalance the flow allocation, as the managed flows are embedded
on the same physical links.
Flow space overlap checks cannot be used to detect
such conflicts.

\stefan{todo: explain figure}

\item \textbf{Example 2: Supporting modifications.} Consider the problem of modifying previously
installed forwarding rules.
While a bundle allows to conditionally add new rules or remove
old rules (and, using the flow space overlap
check feature $\checko$, to abort the transaction if potentially conflicting
rules were already there), it cannot be easily used
to conditionally modify existing rules .
In other words, the bundle and overlap check only help
for the \emph{conditional creation} of rules, not their \emph{modification}.
\end{enumerate}

\begin{figure*}[t]
\centering
\includegraphics[width=1\columnwidth]{no-sync.png}~\includegraphics[width=1\columnwidth]{sync-with-cas.png}\\
\caption{\emph{Left:} No synchronization. \emph{Right:} CAS.}
\end{figure*}

As we will see, we do not even need the bundle feature
to implement the powerful synchronization objects of our transactional interface.
However, the bundle feature can still be useful in the context of
consistent policy updates.

@@

%\section{The Container Object}\label{sec:vision}
\section{Transaction $\&$ Synchronization}\label{sec:main}

Given this motivation, we now present a more powerful transactional interface.
At a high level, we aim to export a
transactional interface which allows each controller $i\in[N]$ ($N\geq
2$) to perform multiple
operations on the configuration of an OpenFlow switch as an atomic
\emph{transaction}, inspired by the transactional-memory paradigm~\cite{stm-st95,tm-book}. A transaction may \emph{commit}, in
which case it appears as executed sequentially, or \emph{abort}, in
which none of its operations takes effect.

@@

\paragraph{The resource claimer}
\petr{
the "resource claimed" part a hard read, the trick of using masks and values is smart, deserves to be explained better
}

>>>>

\newcommand{\error}{\texttt{error}}
\newcommand{\xid}{\texttt{xid}}
\newcommand{\ecode}{\texttt{error-code}}
\newcommand{\bundle}{\textbf{atomic}}
\newcommand{\ack}{\textit{ack}}

@@

\begin{abstract}
Control planes of modern Software-Defined Networks (SDNs)
are \emph{distributed systems}: to ensure availability and fault-tolerance,
to improve load-balancing, and to reduce overheads,
control modules operating on the data plane are physically distributed.
%a redundant control plane
%can ensure a high availability, and a spatially distributed
%control plane allows to handle time-critical dataplane
%events close to their origins.
%[[PK this is a contraversial point, not important for us
%Moreover, an SDN may be managed by multiple administrators simultaneously.
%]]
However, %a distributed control plane also introduces
%new challenges, 
in order to guarantee \emph{consistency} of network operation, actions
performed on the data plane by different controllers may need to be \emph{synchronized},  which is a
highly nontrivial task.
%[[PK sounds contradicting to the fact that such systems already exist
%However, not much is known today about how to implement
%such a distributed control plane.
%]]
%
In this paper, we propose a synchronization
framework for control plane  based on the well-known concept
\emph{transactions}.    
%
A transaction aggregates a sequence  
of control messages and specific synchronization primitives, 
providing all-or-nothing semantics.
We show that a powerful transactional interface can be implemented \emph{in
  band}, using the standard Openflow commands on the data-plane configuration.
%Indeed, recent Openflow versions already come with useful but primitive
%features, in particular \emph{bundling} and \emph{rule-overlap checks}.
%In particular, we present a
%transactional interface which supports arbitrary atomic
%transactions
%over the switch configuration space: 
%Our transactions
%can include complex notions of conflicts (beyond flow overlap checks),
%as well as computations and multiple controller interactions.
We discuss how to use our synchronization framework to implement
control-plane replicated service, consistent policy composition, and load-balancing.
\end{abstract}

@@

While the perspective of \emph{logically centralized} control plane
offered by SDN is intuitive and attractive,
there is a wide consensus that
the control plane should be  \emph{physically distributed}.
First, in order to provide high availability, 
%as well as a minimal degree of
%fault-tolerance, 
%[[PK fault-tolerance is necessary to provide hight availability?]]
controllers should be redundant~\cite{onix,stn,onos}: a failure
of one controller can be masked by other controllers. 
Second, it has also been proposed to distribute controllers \emph{spatially}, in order to handle latency-sensitive and
communication intensive 
%control-plane 
data-plane
events close to their origin~\cite{devoflow,kandoo,jukka,disco}
Third, larger SDNs are likely to be operated by multiple administrators or may even offer
participatory interfaces where different users can install and trigger policy changes
concurrently~\cite{participatory,stn}.

@@

\section{Background: SDN and Openflow}\label{sec:background}
%\petr{ the 1st sentence does not belong here}\liron{I think it serves as a motivation for the section, but it fines by me to remove it}

%In this section, we recall the basic SDN principles, with a focus on
%Openflow~\cite{of-spec}, the standard SDN protocol today.

\paragraph{Control and data planes}
Software-defined networking decouples \emph{control plane}, software
controllers that configure the network, from the \emph{data plane}, the
devices that implement the network.  
% 
In Openflow~\cite{of-spec}, the \emph{de facto} standard SDN protocol
today, the data plane consists of \emph{Openflow switches}. 
In a nutshell, the configuration of an Openflow switch is a set of
\emph{rules}. 
A rule is essentially a \emph{match-action} pair:
the match part of a rule identifies the space of packet headers that are
subject to the rule (e.g., all packets to a specific destination) and
the action part identifies how the switch should process the matched
packets (e.g., forward them to a specific port).

More precisely, the match field of a rule is 
a  a ternary pattern over packet header fields
(a value and a mask).
\petr{example?}
In general, a packet can be matched against any of its header
fields, and fields can be wildcarded and sometimes bitmasked.
\petr{unclear what ``sometimes'' means here}
%(For instance, the meta-data field is maskable.)

A set of rules, associated with \emph{priority} levels, form \emph{flow
  entries} in a \emph{flow table}, and the 
entries are ordered according to their priorities.
Furthermore, an Openflow switch may maintain multiple flow tables
forming a \emph{pipeline}.
Data-plane packets are checked against flow entries in the pipelined
tables to determine the set of actions the switch should perform on the packet.      
A data-plane packet arriving at a switch is first checked against
the rule with the highest priority and in table $0$. 
If the header of the packet fits the match fields of that rule, 
a default \emph{instruction} associates the packet the corresponding action.
Otherwise, the packet is checked against flow  entries with lower
priorities.  
Specific instructions can be used to define
additional tables to be visited (\textit{goto} instructions), to
modify the set of actions associated with the packet (by adding,
deleting or modifying actions), or immediately apply some actions in the action set. 
During the traversal of the flow tables, the packet can also be equipped with \emph{meta-tags} to
transfer temporary information between tables.
The meta-data field is part of the header that can be
inserted or removed from a packet via \emph{push} and \emph{pop}
actions. 
\petr{do we need the discussion of multiple tables/pipelining here? I still find it
  very confusing}

The actions associated with the packet are performed at the end of the pipeline.
If no rule matches the packet, the packet is dropped.

\liron{we should explain that: every packet is associated with an
  action set. This set is empty at the beginning and is changed
  according to the flow table matches. Each match is asociated with
  instructions that can add or modify actions in the action set and
  also to apply (now) all actions in the action set. By default action
  set is applied at the end of the pipeline.}

@@

\paragraph{Interaction between control and data planes: FlowMod commands}
The flow tables installed on the switches across the data plane
determine the network \emph{policy}.
Controllers can change the policy via sending  
control messages containing \emph{FlowMod commands}.
In a nutshell, a \emph{FlowMod} command either specifies a new flow entry or
a modifications to an existing flow entry. \petr{The description below
  does not explain how a flow entry can be modified.}  
A flow entry consists of: (1)~a match 
field
%defined by a ternary pattern over packet header fields
(a value and a mask), (2)~a set of actions, and (3)~a priority.
%,a match and an action.
The standard processing of a FlowMod command received by a switch is
as follows. 
If the switch already maintains a flow entry with \emph{exactly} the
same match, then the new flow entry will simply \emph{replace} it.
\petr{What if the two rules have different priorities?}
Otherwise, the flow entry will be installed in addition to existing ones. 
%but will be installed
%\emph{in addition} to existing entries if they only
%partially overlap (or define flowspace subsets or supersets).

If rules with overlapping but not identical match fields have the same priority, 
inconsistencies can arise, since there is no clearly specified action
that should be applied to a packet matching the two entries.
In order to avoid such conflicts, a \emph{FlowMod} command can be equipped with a the $\checko$ flag:
if the switch maintains is a flow entry with an overlapping but not
identical match part with the same priority but a different action,
 and if the two rules have the same priority, then
\emph{FlowMod} will fail.
\petr{Is ``different action'' important here?}
The $\checko$ flag in a \emph{FlowMod} command helps resolving simple
conflicts between controllers, and will be instrumental
in implementing our synchronization abstractions.

%[[PK not sure this is the right place 
%implement synchronization objects which form
%the basis of our transactional interface.
%]]
We will use the following notation to create a \emph{FlowMod} message:
$\FlowMod(\match, \action, \flags)$, $\match$ denotes the match,
$\action$ the action, and $\flags$ the flags to activate the $\checko$
feature.
\petr{Sometimes we also write something like op=add in flowmod, should
  we specify it here?}  
The \emph{FlowMod} message (and also the flow entries) include more parameters,
which however are irrelevant for our implementation: we suggest to
use the default values.

By installing Openflow \emph{flow monitors} with the \textsf{OFPFMF\_INITIAL} optional flag,  
a controller can read the current configuration. 
% %(and subscribe for updates). 
%
%can i match multiple flows in multiple tables? yes, but only for FlowMod, not GroupMod: but that is good enough!%
\liron{I changed a bit the description of flow monitors following Petr's comment, I hope it's more clear now.}
\petr{Not very clear what reading the configuration means: is it done
  in an atomic way?}

%[[PK should go to the discussion?
%We also remark that our solution is only based on flow tables
%and does not make use
%of \emph{Openflow group tables}, nor do we make use of \emph{cookies}: rules
%are always identified by exact matches.
%]]

@@

Another important feature in Openflow is the \emph{bundle}. 
A controller can send multiple $\FlowMod$ commands  equipped with 
the same \emph{bundle identifier}.
These commands are buffered temporarily by
the switch until a \emph{bundle commit} message is received from the
controller. 
The bundled commands are then performed with \emph{all-or-nothing semantics}:
either all of the commands are performed, or none of them is. 
In particular, if a configuration request contained in one of the bundled commands cannot
be applied (e.g.,  all the command is rejected because of the set
$\checko$ flag and a conflicting flow entry), all  other commands in
the bundle are rejected and an error
message is sent to the controller that issued them.
%If multiple \emph{FlowMod} messages with the $\checko$ flag are sent as a single bundle,
The error message contains $\xid$, the identifier  first failed command in the
bundle, and $ecode$, the \emph{error code} corresponding to the failure.   
%[[PK too detailed and not very clear
%Concretely, when the commit fails, the switch must generate an error message corresponding to the
%message that could not be applied. The error message must use the identifier (\emph{xid}) of the offending message,
%the error data field corresponding to that message and the error code corresponding to the failure. If
%multiple messages are generating errors, the switch may return only the first error found or generate
%multiple error messages for the same bundle.
%]]

When properly configured,
a bundle can be processed in a \emph{packet-atomic} way, in the sense that
each packet is processed according
to the switch configuration either \emph{before} or \emph{after} the
bundle (i.e., all its commands) are executed. 
\liron{we need to disscuss controller atomicity and order preservation inside a bundle}
\petr{Right! :)}

@@

@@

\section{Limitations and Motivations}\label{sec:motivation}

% generic way for multiple distributed controllers to safely update switch configuration with no need of any control plane

At first sight, bundling \emph{FlowMod} commands with  the $\checko$
flag already provides a very powerful synchronization mechanism.
Indeed, its ``min-transaction'' nature allows a controller 
to install multiple flow entries in an atomic way, making
%[[PK we did not say what reads and writes are... 
%At first sight, the Openflow bundle feature seems to provide everything
%we need to implement our transactional interface \petr{did we say what we need already?}: the bundle
% defines a sequence of read and write operations
% on the switch state which is executed atomically.
%]]
However, as is 
%while the bundle can indeed be seen as a ``mini-transaction'', 
bundling has important limitations.
%\petr{unclear what "reads and writes" are here, not discussed before. How openflow can be used to implement reads and writes?}

First, %if employed without any additional features, 
a bundle does not provide any means to modify the switch configuration 
based on a generic application-specific condition, depending on the
current configuration, except to limited types of constraints, e.g.,
or \emph{overlapping flow spaces} or insufficient free space.
Indeed, sometimes it is desirable to  define more generic conditions
under which configuration updates should or should not take place.

%But also bundling FlowMod commands with the check overlap feature ($\checko$)
%comes with limitations: while it is possible
% to \emph{abort} a bundle depending on certain conditions
% in the \emph{overlapping
% flow spaces}, oftentimes it is desirable to be able
% to define more general conditions and notions of conflicts
% under which an atomic update should not take effect.
% \liron{replaced transaction in the last sentence with " atomic update"}

Second, sometimes it is desirable to react to conflicts in smarter ways than
 to simply abort, e.g., by supporting \emph{conditional modifications}.
Let us consider two examples.

\begin{enumerate}
\item \textbf{Example 1: Handling complex dependencies.} Let us consider two controllers
in charge
of \emph{load-balancing}.
In a load-balancing application, seemingly independent actions, defined over
completely independent logical flow spaces (say, two flows with source IP address ranges $1*$ and $0*$),
may in fact be dependent due to collocation in the underlying physical network.
If the two controllers concurrently and independently
update forwarding rules,
they may involuntarily
unbalance the flow allocation, as the managed flows are embedded
on the same physical links.
Flow space overlap checks cannot be used to detect
such conflicts.

Consider the example in the left part of Figure~\ref{fig:CAS-example}. Two controllers,
$1$ and $2$ read the configuration, try add two more flows to a switch
with two outgoing links, balancing the load.  
Imagine that each of the controllers reads the switch configuration and
learns that currently one flow is forwarded to the left link and two
flows are forwarded to the right link.
Without synchronization, the two controllers naturally choose to
install their flows on the left link, which results in an undesirable
unbalanced state. Note that, assuming that all these flows are
completely independent, bundling and $\checko$ features do not really
help here.

\petr{postpone the explanation of the right part of the figure until we come to CAS?}

\item \textbf{Example 2: Supporting modifications.} 
Consider the problem of modifying previously installed forwarding rules.
While a bundle allows to conditionally add new rules or remove
old rules (and, using the flow space overlap
check feature $\checko$, to abort the transaction if potentially conflicting
rules were already there), it is not straightforward how to use it to  
modify existing rules, especially provided a specific condition.
In other words, the bundle and overlap check only help
for the \emph{conditional creation} of rules, not their \emph{modification}.
\end{enumerate}

Intuitively, one way to address these issues is to use a part of the
configuration space of the data-plane to store 
``meta-configuration'', i.e., information on contention and conflicts between
concurrently installed policy updates. Controllers can read and modify
this meta-data by applying synchronization primitives. 
In the following section, we
describe how this space can be organized and consistently maintained
by multiple controllers despite of concurrency and conflicts.

\begin{figure*}[t]
\centering
\includegraphics[width=1\columnwidth]{no-sync.png}~\includegraphics[width=1\columnwidth]{sync-with-cas.png}\\
\caption{\emph{Left:} No synchronization. \emph{Right:}
  CAS.}\label{fig:CAS-example}
\end{figure*}

%[[PK not really discussed?
%As we will see, we do not even need the bundle feature
%to implement the powerful synchronization objects of our transactional interface.
%However, the bundle feature can still be useful in the context of
%consistent policy updates.

@@

%\section{The Container Object}\label{sec:vision}

\section{Synchronization abstractions}\label{sec:main}

We now describe how our  meta-configuration space at
an Openflow switch can be implemented and manipulated using our
synchronization framework.

\subsection{Configuration space}

We distinguish two parts of the configuration space of a switch:
the ``normal'' configuration which determines the network policy and
is used to process the data-plane traffic (e.g., forwarding rules), 
and the ``meta'' configuration that contains information used by the
controllers to synchronize their actions.

\petr{How to make sure that traffic is not affected by the ``meta'' part?}

\subsection{Transactional interface}

Our framework provides the controllers with a transactional interface to the
switch configuration.     
A transaction consists of a sequence of \emph{operations}, which
should either be executed in its entirety or completely rejected. 
An operation can be a regular FlowMod command or a synchronization
primitive, which, as we
show below translate into specific FlowMod commands operating on the
switch meta-configuration.

We will use the notation
%$\textbf{bundle}\{\;\ldots\;\}\rightarrow res$ to refer to represent a
$\bundle\{op_1,\;\ldots\;op_k\}$ to refer to a
sequence of commands $op_1,\ldots,op_k$ sent in a bundle.
%\petr{What about replacing $\top-\bot$ with ack-nack? Can we also get a reason why a bundle did not commit?}
The bundle construct is implemented as a sequence of the following
control messages:
\begin{enumerate}
\item A \emph{create bundle} message: a \texttt{ofpt\_bundle\_control} message of type
\texttt{ofpbct\_open\_request} equipped with \textit{bid}, the bundle identifier.
\item A sequence of commands, each wrapped in an
\emph{add to bundle message} (\texttt{ofpt\_bundle\_add\_message})
equipped with \emph{bid}.
\item Finally, a \emph{commit bundle} control message (\texttt{ofpt\_bundle\_control}) of
type \texttt{ofpbct\_commit\_request} with \textit{bid}.
\end{enumerate}
The construct returns $\ack$ if the bundle \emph{commits} and all commands are performed successfully, and
$\error(\xid,\ecode)$ received from the switch otherwise.
%Again, the bundle always returns a binary result: \emph{error} ($\bot$) or \emph{success} ($\top$).
\petr{Should also provide the reason why the bundle has failed?}

\hide{
Given this motivation, we now present a more powerful transactional interface.
At a high level, we aim to export a
transactional interface which allows each controller $i\in[N]$ ($N\geq
2$) to perform multiple
operations on the configuration of an OpenFlow switch as an atomic
\emph{transaction}, inspired by the transactional-memory paradigm~\cite{stm-st95,tm-book}. A transaction may \emph{commit}, in
which case it appears as executed sequentially, or \emph{abort}, in
which none of its operations takes effect.
}

\subsection{Synchronization primitives}

@@

\paragraph{The resource claimer}
\petr{
the "resource claimed" part a hard read, the trick of using s and values is smart, deserves to be explained better
}
